An event-sourced and decentralized database with IPLD and materialized views

### 2019-09-23 draft.tex
fixes alignment
Signed-off-by: andrew <andrew@textile.io>
update: rough first draft
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

\subsection{Thread Views} \label{sec:threadviews}

The use of well-established design patterns from the CQRS+ES (and related) literature means the Threads protocol can provide the foundation for many higher-level concepts and technologies. Like many CQRS-based systems, the ``public'' API for Threads revolves around the concept of Views, which are used to optimize queries on the local event store. A key feature of Views is they are completely disposable, because they can be entirely rebuilt from the source event store\footnote{Though see section \ref{sec:SnapshotsCompaction} for a discussion of Snapshots, which may be used to speed up Views.}. In Threads, Views are updated by subscribing to specific Event(s), and applying a Reducer function that is used to update the View state (which may be \emph{materialized} to local persistent storage). This simple subscription system provides a flexible framework for building complex View-driven application logic.

\subsubsection{Databse Views}\label{sec:DefaultViews}

The public API for a Thread already provides several features that you would expect when operating on a dataset or table within a database. Indeed, each Thread is defined by a unique ID (its Identity), and provides facilities for access control and permissions, networking, sharing, invites, and more. As such, it is not difficult to imagine a simplified API in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.

@@

A database built on Threads would map database operations into Events. These Events would then mutate a View Model via its corresponding Reducer function, as per above. The actual View could be persisted or kept in-memory, and would be directly queried via a corresponding database-style APIs. Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} Event, marking the database item for removal from the underlying materialized View.

@@

A key/value store built on Threads would ``map'' these database Events into key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del}. These Events would then mutate a map-like View Model. To complete the database-like API, the entire API could be encapsulated in a database sturcture that satisfies a given interface (see for example, Figure \ref{ex:KVStore}).

@@

Eventually consistent, CRDT-based stores can also be implemented on top of Threads. CRDT-based stores are particularly useful for managing Views of a document in a multi-peer collaborative editing environment (like Google Docs or similar). For example to support offline-first, potentially concurrent edits on a shared JSON document, one could implement a JSON CRDT datatype \cite{kleppmannConflictFreeReplicatedJSON2017} that applies updates on a JSON document View model. Libraries such as Automerge\footnote{\url{https://github.com/automerge/automerge}} provide useful examples of reducer function that make working with JSON CRDTs relatively straightforward, and implementations in other programming languages are also available. A practical example of using a JSON CRDT in Textile is given in section \ref{sec:AccessControl}, where it is used to represent updates to an ACL document for an access-controlled Thread implementation.

\paragraph{Application State}

For application developers familiar with common state management issues, tools such as Redux provide an opinionated framework for making state mutations predictable by imposing certain restrictions on how and when updates can happen. In fact, Redux builds on concepts from CRQS and ES itself \cite{reduxMotivation}, and combines them with three core principals:

\begin{enumerate}
\item The entire \emph{state} of an application should be stored in a single \emph{store}.
\item The only way to change the state is to emit an \emph{action} (or event).
\item transforms the application state using a \emph{pure} (i.e., zero side effects) reducer function (projection).
\end{enumerate}

In practice, Redux doesn't care how a developer persists the application state, which means a Thread's event store and read-only materialized views are excellent candidates. This focus on unidirectional data flow in perfectly in line with architecture proposed in this paper, and as such, a compatible Redux API will be built on top of Textile Threads.

@@

At an abstract level, the Textile protocol provides a highly distributed framework for building shared, offline first, datastores that are fault tolerant\footnote{When using an ACID compliant backing store for example.}, eventually consistent, and scalable. Any internal implementation details of a compliant Threads ``client'' may use any number of well-established design patterns from the CQRS+ES (and related) literature to ``extend'' the Threads protocol with additional features and controls. Some examples of extensions that are included by default in Textile's Threads implementation are outlined in this section to provide some indication of the extensibility of Threads.

@@

Snapshots\footnote{The literature around snapshots and other CQRS and ES is somewhat confusing, we attempt to use the most common definitions here.} are simply the current state of a View at a given point in time. They can be used to rebuild the state of a view without having to query and re-play all previous events. When a Snapshot is available, a Thread Peer can rebuild the state of a given View by replaying any events generated since the latest Snapshot (with a corresponding reduction in the total number of events that need to be processed) using the View's Reducer function. Multiple peers processing the same Log could create a Snapshot every 1000 events and be guaranteed to create the exact same Snapshot because each Peer's Event counts are identical\footnote{Assuming any network partitions are only short-lived (i.e., that peers are able to share events consistently).}.

In practice, Snapshots are written to their own Event log and stored locally. They can potentially be synced (Section \ref{sec:LogSync}) to other peers as a form of data backup or to optimize state initialization when a new peer starts participating in a shared Thread (saving disk space, bandwidth, and time). They can similarly be used for initializing state during recovery.

Compaction can similarly speed up re-hydration of state, and is useful when only the latest Event per Event type is required. In addition to being used as a starting point for initialization and recovery, compacted logs can be used to \emph{replace} the Log from which they were created in order to free up local disk space. In Textile, compaction becomes much easier because a log only has one writer and conflicts are not possible. In practice, Log compaction is a local-only operation (i.e., other Peers do not need to be aware that Compaction was performed).

@@

One of the most important properties of a shared data model is the ability to apply access control rules. There are two forms of access control possible in Threads, Sequence-level ACLs and Thread-level ACLs. Thread-level access control lists (ACLs) allow creators to specify who can \emph{follow, read, write, and delete} Thread data. Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (see Definition \ref{ex:SequenceID}) basis. Both types of ACLs are materialized views, or more specifically, JSON CRDT documents, built from Log Events (Section \ref{sec:TexCRDT}). ACLs implemented as JSON documents provides two advantages over static or external ACL rules (although static and external ACLs are also possible). First, ACLs are fully mutable, allowing developers to create advanced rules for collaboration with any combination of readers, writers, and followers. Second, because ACLs are JSON documents, ACLs can specify their own editing rules (i.e. allowing multiple Thread participants to modify the ACL) in a self-referencing way.

>>>>

\section{Thread Internals}

To make Threads as easy to adopt and use as possible, we've designed the internal event store to take advantage of ideas from several existing CQRS and ES systems, as well as ideas and designs from Flux\footnote{http://facebook.github.io/flux}/Redux\footnote{Flux/Redux build on concepts from CRQS and ES themselves, and Redux is arguably an implementation of the Flux application architecture (i.e., Flux is a pattern, Redux is a library).} \cite{reduxMotivation} and domain driven design [\dots] (DDD). In the following sections, we outline the internal structure of a local Threads ``client'' at a relatively high-level, and outline how it can be used to build intuitive developer-facing APIs to make adopting and using Threads ``the right choice'' for a wide range of developers.

\subsection{Overview}

For application developers familiar with common application state management issues, tools such as Flux/Redux provide opinionated frameworks for making state mutations \emph{predictable} by imposing certain restrictions on how and when updates can happen. If you take a look at the architecture of a Flux (or Redux) application, you'll notice several core similaries to CQRS and ES systems (see Figure \dots). In both cases, the focus is on unidirectional data flows that build downstream views from atomic updates in the form of events (or actions).

We adopt a similar flow in Threads, such that views are updated by subscribing to specific events (or actions as in Flux/Redux), and applying a reducer function that is used to update the view state (which may be \emph{materialized} to local persistent storage). This simple subscription system provides a flexible framework for building complex event-driven application logic. In the remdainder of this section, we outline the specifics of the Textile Threads internals, focusing on three core components: Actions, the Dispatcher, and the Read/View Stores.

\subsubsection{Actions}

Like with Flux/Redux, to add an Event to the internal Threads system, we use \emph{Actions}. Actions cause predictable updates to state via \emph{Reducer} functions. In practice, Actions are \emph{dispatched} from a singleton \emph{Dispatcher}, which then stores the Event in an \emph{Event Store}, and calls a set of pure (in the functional sense) Reducer functions that mutate a set of downstream \emph{Read/View Stores}.

In addition to dispatching 'raw' Actions, application developers are able to build and interact with ction \emph{Creators}, which are simple functions (or groups of functions) that create (and possibly dispatch) new Actions. For those familiar with DDD, if built around a specific \emph{domain}, Action Creators provide bounded context that can be (roughly) compared to an aggregate root (i.e., a cluster of objects, that create a clear reference to the root aggregate). This is a powerful design feature that we will take advantage of in section \dots.

\subsubsection{Dispatcher}

In order to actually \emph{store} events and dispatch actions to downstream views, a Dispatcher is used. The dispatcher wraps an embedded Event Store, which is responsible for persisting Events. This means a Dispatcher is both an Event Store an Event Bus/Dispatcher. In practice, the internal Event Store supports ACID transactions [\dots], so that Actions are only ever dispatched if they were successfully persisted to the Event Store.

All ``side effects'' of a given Action happen within a `Transaction`, in order to ensure consistently of both storage (Event Store) and actions. Transactions provide a similar experience to Redux ``Sagas'' [\dots], but with a simpler interface. Once an Event has been persisted in the internal Event Store, the Dispatcher is responsible for running a `Store`'s `Reducer` callback. In order to make this possible, Stores must \emph{registered} with the Dispatcher.

Actions can also be batch dispatched, which is the preferred way to dispatch multiple Actions at a time. This works by grouping Actions into meta Actions, and then running their Reducers in series. An example of where this might be useful is when grouping Actions that regularly occur together, such as image processing pipelines.

\subsubsection{Read Models}

The Dispatcher, combined with Actions and Creators provide a useful interface for the write side of the Threads CQRS framework. Like many CQRS-based systems, the read side revolves around the concept of views. Here, we borrow some ideas and terminology from a combination of Flux/Redux and CQRS, but with a slightly more abstracted API layer, similar to DDD.

As Actions as dispatched on the write side, they are reacted to on the read side via Reducers, causing updates to Read Models, which provide interfaces for queries and accessing state. In practice, a ReadStore is defined by a Reducer function and a set of query Resolver functions.

The Reducer functions build a ReadStore's state based on incoming Actions. Query Resolvers use data from the accumulated state to answer specific queries. Reducers should be ``pure'' functions (where possible)\footnote{In practice they aren't really pure, because we are mutating (persistent) state here}. This is very similar to a Flux Store, and it is possible to have multiple Stores here, each specifying their own model and Reducer function.

Similarly, Threads defines a ViewStore, which are a special kind of ReadStore, akin to a Redux Store. In practice, they are smaller, read-only (immutable) models, that are built on-the-fly, and defined by a pure Reducer function. Threads use ViewStores to define the ACL for a given Thread.

Like in Flux, Stores (as they are outlined above) have several specific properties:

\begin{itemize}
  \item Cache data
  \item Expose public getters to access data (never have public setters)
  \item Respond to specific `Actions` from the `Dispatcher`
  \item Always emit a change when their data changes
  \item Only emit changes during a `Dispatch`
\end{itemize}

To query Stores, we use their set of Resolver methods. These are just pre-defined query functions\footnote{Ideally the underlying storage is optimized to support these queries} that exist on the Store instance (the public APIs so to speak). The interesting thing here is that these can similarly be defined around a domain to produce essentially domain aggregates in the DDD sense.

\subsection{Abstractions}\label{sec:DefaultViews}

Building on the idea of bounded context and aggregate roots from DDD, Threads provides several default abstractions based on the idea that a database provides a useful context within which to structure our Events/Actions. Indeed, the components of a Thread already provide several features that you would expect when operating on a dataset, or table within a database: each Thread is defined by a unique ID, and provides facilities for access control and permissions, networking, sharing, invites, and more. As such, it is not difficult to imagine a simplified API in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems --- database domains.

@@

A database built on Threads would map database operations into \texttt{Actions}. Here, the \emph{domain} is the databse. These \texttt{Actions} would then mutate a \texttt{ReadModel} via its corresponding \texttt{Reducer} function, as per above. The \texttt{Store} could be persisted or kept in-memory (as a \texttt{ViewModel}), and would be directly queried via corresponding database-style \texttt{Resolver} metods. Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} \texttt{Action}, marking the database item for removal from the \texttt{ReadStore}.

@@

A key/value store built on Threads would ``map'' these database \texttt{Actions} into key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del}. These \texttt{Actions} would then mutate a map-like \texttt{ReadModel}. To complete the database-like API, the \texttt{ActionCreator} and \texttt{ReadModel} methods outlined in Section \dots could be encapsulated in a database structure that satisfies a given interface (see for example, Figure \ref{ex:KVStore}).

@@

Eventually consistent, CRDT-based stores can also be implemented on top of Threads' Event-driven architecture. CRDT-based Stores are particularly useful for managing views of a document in a multi-peer collaborative editing environment (like Google Docs or similar). For example to support offline-first, potentially concurrent edits on a shared JSON document, one could implement a JSON CRDT datatype \cite{kleppmannConflictFreeReplicatedJSON2017} that merges updates to a JSON document in a ReadModels Reduder function. Libraries such as Automerge\footnote{\url{https://github.com/automerge/automerge}} provide useful examples of reducer functions that make working with JSON CRDTs relatively straightforward, and implementations in other programming languages are also available [\dots]. A practical example of using a JSON CRDT in Threads is given in section \ref{sec:AccessControl}, where it is used to represent updates to an ACL document defined as a ReadModel, with queries defined for an access-controlled Thread implementation.

@@

At an abstract level, the Textile protocol provides a highly distributed framework for building shared, offline first, Stores that are fault tolerant\footnote{When using an ACID compliant backing store for example.}, eventually consistent, and scalable. Any internal implementation details of a compliant Threads ``client'' may use any number of well-established design patterns from the CQRS and ES (and related) literature to ``extend'' the Threads protocol with additional features and controls. Indeed, by designing our system around Actions, a Dispartcher, and generic ReadStores, we aim to make easy to extend Threads in many different ways. Some examples of extensions that are included by default in Textile's Threads implementation are outlined in this section to provide some indication of the extensibility this design affords.

@@

Snapshots\footnote{The literature around snapshots and other CQRS and ES is somewhat confusing, we attempt to use the most common definitions here.} are simply the current state of a Store at a given point in time. They can be used to rebuild the state of a ViewStore without having to query and re-play all previous events. When a Snapshot is available, a Thread Peer can rebuild the state of a given ViewStore by replaying any events generated since the latest Snapshot (with a corresponding reduction in the total number of events that need to be processed) using the Stores's Reducer function. Multiple peers processing the same Log could create a Snapshot every 1000 events and be guaranteed to create the exact same Snapshot because each Peer's Event counts are identical\footnote{Assuming any network partitions are only short-lived (i.e., that peers are able to share events consistently).}.

In practice, Snapshots are written to their own Event log and stored locally. They can potentially be synced (Section \ref{sec:LogSync}) to other peers as a form of data backup or to optimize state initialization when a new peer starts participating in a shared Thread (saving disk space, bandwidth, and time). They can similarly be used for initializing a Store's during recovery.

Compaction can similarly speed up re-hydration of a Stores's state, and is useful when only the latest Action of a given type is required. In addition to being used as a starting point for initialization and recovery, compacted Stores can be used to \emph{replace} the Store from which they were created in order to free up local disk space. In practice, Compaction is a local-only operation (i.e., other Peers do not need to be aware that Compaction was performed).

@@

One of the most important properties of a shared data model is the ability to apply access control rules. There are two forms of access control possible in Threads, Sequence-level ACLs and Thread-level ACLs. Thread-level access control lists (ACLs) allow creators to specify who can \emph{follow, read, write, and delete} Thread data. Similarly, sequence-level ACLs provide more granular control to Thread-writers on a per-sequence (see Definition \ref{ex:SequenceID}) basis. Both types of ACLs are implemented as JSON CRDT ViewStores (see Section \ref{sec:TexCRDT}). ACLs implemented as JSON ViewStores provide two advantages over static or external ACL rules (although static and external ACLs are also possible). First, ACLs are fully mutable, allowing developers to create advanced rules for collaboration with any combination of readers, writers, and followers. Second, because ACLs are JSON documents, ACLs can specify their own editing rules (i.e. allowing multiple Thread participants to modify the ACL) in a self-referencing way.
